Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-05 tagless/tagless-final.tex
small changes to the contributions

Our code ensures statically that an object term is well-typed and
closed, yet is surprisingly simple and obvious in hindsight.  We stress
that until now it has been considered impossible to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
states that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We further discuss related work in~\S\ref{related}.

Our contributions are
\begin{enumerate}
\item to evaluate, compile, and partially evaluate a typed object language
   in a typed metalanguage with a simple type system;
\item to check statically for well-typed and closed object programs;
\item clean, comparable implementations in MetaOCaml and Haskell;
\item a functor signature that encompasses all our interpreters, from
    evaluation and compilation to partial evaluation and CPS transformation;
\item a clear way to extend the object language with more features;
\item and an approach to self\hyp interpretation compatible with the
  above.  Self\hyp interpretation turned out to be harder than expected.
\end{enumerate}

\jacques{we should probably be more explicit about the contributions of
the discussion section}

@@

In the sequel, we encode binding in object programs using higher-order
abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices.

@@

To embed the same object language in (Meta)OCaml, we replace the type
class |Symantics| and its instances by a module signature |Symantics|
and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
signature that suffices for this and the following sections.
Two differences between this signature and the |Symantics| class in
Haskell are: the additional type parameter |c|, an
environment classifier \citep{WalidPOPL03} required by MetaOCaml for
code generation in~\S\ref{compiler}; and the $\eta$-expanded type for
|fix| and thunk types in |if_| due to OCaml's being a call-by-value
language.

>>>>

Our code ensures statically that an object term is well-typed and
closed, yet is surprisingly simple and obvious in hindsight.  We stress
that until now it has been considered impossible to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
state that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We further discuss related work in~\S\ref{related}.

Our contributions are
\begin{enumerate}
\item to evaluate in direct or continuation-passing styles, 
  compile, and partially evaluate a typed higher-order object language
   in a typed metalanguage with a simple type system. All these
   interpreters have no type tags and never get stuck;
\item a Hindley-Milner--typed partial evaluator with no 
  polymorphic lift operation and with
  the delayed binding-time analysis of functions . The partial
  evaluator needs no advanced type systems such as GADT or intentional
  type analysis;
\item to check statically for well-typed and closed object programs
      using the type system \oleg{typechecker?} of the metalanguage;
\item clean, comparable implementations in MetaOCaml and Haskell;
\item a functor signature that encompasses all our interpreters, from
    evaluation and compilation to partial evaluation and CPS;
\item a clear way to extend the object language with more features
  such as state;
\item and an approach to self\hyp interpretation compatible with the
  above.  Self\hyp interpretation turned out to be harder than expected.
\end{enumerate}

@@

In contrast to \S\ref{tagproblem}, we encode 
binding in object programs using higher-order
abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices. This makes the encoding convenient and
also ensures that our object programs are closed.

@@

To embed the same object language in (Meta)OCaml, we replace the type
class |Symantics| and its instances by a module signature |Symantics|
and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
signature that suffices until \S\ref{PE}.
Two differences between this signature and the |Symantics| class in
Haskell are: the additional type parameter |c|, an
environment classifier \citep{WalidPOPL03} required by MetaOCaml for
code generation in~\S\ref{compiler}; and the $\eta$-expanded type for
|fix| and thunk types in |if_| due to OCaml's being a call-by-value
language.
